The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens. Or Identifying when environmentally persistent pathogens transform good habitat into ecological traps

### 2015-06-30 Manuscript/manuscript.tex
Initial edits distinguishing metapop vs. ind level traps

We expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Robertson2006}, \cite{Almberg2011}).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).  Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.

@@

where $\phi(X, t)_j$ is an indicator function that is $1$ if patch $j$ is in state $X$ at time $t$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the effective distance between patches $i$ and $j$ .  Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.

@@

The model was parameterized  according to the following assumptions (Table~\ref{params}). (1) The rate of emigration from a patch increases with population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).  (2) Migrants preferentially select high quality habitat, that is, high quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}, though we relax this below).  (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$).  To explore the dynamics of a range of possible pathogens, disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity (representing the half-life of the pathogen's infectivity in the environment) were varied over a range of possible values.

@@

To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate management actions that add either low or high quality habitat respectively (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).  
We then implemented a full-factorial design on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).  In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).

@@

To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).  To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).  We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk and the corresponding fitness costs) and the final total effective host population size.

@@

For a wide range of parameter values, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions despite generally facilitating pathogen spread (red shaded regions in Fig~\ref{highvlow}).  (PCC: while true, one could also say: for the cases where the host and pathogen both persist there was a wide range of parameter values where low quality habitats supported more hosts.  Essentially, I'm intersecting Fig. 1 and 2, excluding the regions of the parameter space where the host or pathogen goes extinct (which is most of the red region in Fig 2). This is a very different sort of take home message. This is just a change of spin compared to how you phrase the paragraph.)
In the case of pathogens that were unable to invade due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}). 
Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e. at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).

However, for pathogens falling between these extremes (e.g. with intermediate disease-induced mortality), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}). (PCC: Yes, may want to lead with this.)
The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
In these cases, the more widespread disease-induced mortality faced by hosts on high quality habitat distributions results in smaller total population sizes than on low quality habitat.
As pathogen longevity increases, this range of pathogens for which low quality habitat is beneficial to the host population becomes smaller, with this benefit apparent only for the most virulent pathogens.
Again, metapopulations with a lattice structure displayed similar patterns (Supplement).

To better understand the mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range (again with $\nu = 0.2$ and $\delta = 0.3$).  
Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  
However, disease risk, measured by the proportion of time a patch spent occupied by infectious hosts, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).  
Thus, from the perspective of a migrant individual, colonizing high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e. $\xi_{im} = 0.5$) increasing this disease risk further (Fig~\ref{pinfection}, a-c vs. d-f).
Moreover, preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing extinction risk substantially compared to random habitat selection (i.e. $\xi_{em} = 0$, Fig~\ref{pext}).

>>>>

We expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Robertson2006}, \cite{Almberg2011}).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).  Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.
The combination of these effects may then lead to the creation of metapopulation-scale traps, where the inclusion of otherwise high quality habitat in the metapopulation actually reduces host population size relative to low quality habitat.

@@

where $\phi(X, t)_j$ is an indicator function that is $1$ if patch $j$ is in state $X$ at time $t$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of the host's mean dispersal distance; and $d_{ij}$ is the effective distance between patches $i$ and $j$ .  Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.

@@

The model was parameterized  according to the following assumptions (Table~\ref{params}). 
(1) The rate of emigration from a patch increases with population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).  
(2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{em}=0.5$, \cite{Hanski2003}, though we relax this below).  
(3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  
In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$).  To explore the dynamics of a range of possible pathogens, disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity (representing the half-life of the pathogen's infectivity in the environment) were varied over a range of possible values.

@@

To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate management actions that add either low or high quality habitat respectively (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).  
We then implemented a full-factorial set of simulations on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).  In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).

@@

To more carefully explore the effects of habitat quality on patch-level dynamics, we performed an additional simulation experiment on metapopulations with patch qualities evenly spaced over the full range explored above (from 0.2 to 1.8).
In these simulations, we varied longevity over the values used above and explored the consequences of host preference for high quality habitat (and the corresponding potential for ecological traps) by choosing two different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).  
To isolate the effects of habitat preference and its interaction with pathogen longevity, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), and landscape structure (fully connected).
We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by susceptible and infectious hosts, and the final total host population size.

@@

For a wide range of parameter values, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions despite generally facilitating pathogen spread (red shaded regions in Fig~\ref{highvlow}).  
In the case of pathogens that were unable to invade due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}). 
Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e. at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).

However, for pathogens falling between these extremes (e.g. with intermediate disease-induced mortality), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
In these cases, high quality habitat serves as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the host population in the presence of disease.
As pathogen longevity increases, the range of pathogens for which we observe this metapopulation-level trap becomes smaller, apparent only for the most virulent pathogens.
Again, metapopulations with a lattice structure displayed similar patterns (Supplement).

To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range (again with $\nu = 0.2$ and $\delta = 0.3$).  
Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  
However, disease risk, measured by the proportion of time a patch spent occupied by infectious hosts, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).  
Thus, from the perspective of a migrant individual, colonizing high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e. $\xi_{im} = 0.5$) increasing this disease risk further (Fig~\ref{pinfection}, a-c vs. d-f).
Moreover, preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing extinction risk substantially compared to random habitat selection (i.e. $\xi_{em} = 0$, Fig~\ref{pext}).
